---
title: "Select variables to plot a heatmap"
author: "`r if (exists('yml')) yml$analyst else 'Jim Zhang'`"
date: "`r Sys.Date()`"
output:
  html_document:
    number_sections: yes
    self_contained: no
    toc: yes
    toc_float:
      collapsed: no
---


<div style="border:black 1px solid; padding: 0.5cm 0.5cm">
**Introduction** This analysis runs a simple procedure that identifies variables significantly different across sample groups via ANOVA and then plots a heatmap of these variables.
</div>

&nbsp;

```{r global_setup, eval=TRUE, include=FALSE}
name.yaml <- 'plot_heat.yaml'; 
name.packages <- c('rmarkdown', 'knitr', 'kableExtra', 'yaml', 'DT', 'htmlwidgets', 'gplots', 'RoCA', 'awsomics'); 
name.subfolders <- c('path.input'='input', 'path.r'='R', 'path.fig'='figure', 'path.tbl'='table', 'path.cl'='cluster'); 

## Default knitr parameters
knitr::opts_chunk$set(dpi=300, dev=c('png', 'pdf'), echo=FALSE, warning=FALSE, message=FALSE, fig.path = 'figure/');
if (!require(devtools)) { install.packages('devtools'); require(devtools); }
if (!require(RCurl)) { install.packages('RCurl'); require(RCurl); }
if (!require(RoCA)) { install_github('zhezhangsh/RoCAR'); require(RoCA); }

## Load required R packages
loaded <- LoadPackage(name.packages);
if (length(loaded[!loaded])) stop('Error: failed to load package(s) ', paste(names(loaded[!loaded]), collapse=', '));

## By default, before knitting this R Markdown file, the YAML file pairing it has been loaded.
## But, the developer can also provide the option to load the YAML file on the fly.
## So, the template can be run using the "Knit HTML" button in RStudio
if (!exists('yml'))                   # if the 'yml' variable doesn't exist yet, create it by loading the YAML file
  if (file.exists(name.yaml))         # assume the pairing YAML file exists in the current folder with the same name
    yml<-yaml.load_file(name.yaml);   # rename the YAML file to fit this template

prms <- yml$parameter;

## Generate directory and sub-directories where the output files will be
f <- GenerateFolder(yml$output, name.subfolders);
path <- yml$output;
for (i in 1:length(name.subfolders)) assign(names(name.subfolders)[i], f[name.subfolders[i]]); 

## Automatical figure/table ordering
OrderFigure(reset=TRUE);
OrderTable(reset=TRUE);

## URL to project home
## Use this line to add a link to project home in the report: `r home.url`
home.url <- Link2Home(yml$home);
```

```{r load_data, include=FALSE}
## Load input data
mtr <- as.matrix(ImportTable(DownloadFile(yml$input$data, path.input))); # tab-delimited data matrix
grp <- ImportVector(DownloadFile(yml$input$group, path.input));

if (length(grp) != ncol(mtr)) stop('Error: column number of data matrix not equal to number of samples\n');
if (length(unique(grp)) < 2) stop('Error: less than 2 sample groups, cannot run ANOVA\n');
if (length(unique(grp)) == length(grp)) stop('Error: none of the sample groups has replicates, cannot run ANOVA\n'); 
names(grp) <- colnames(mtr); 

saveRDS(mtr, paste(path.r, 'data.rds', sep='/')); 
saveRDS(grp, paste(path.r, 'group.rds', sep='/')); 

# re-order matrix column, so samples of the same group are next to each other
ind <- lapply(unique(grp), function(u) which(grp==u)); 
mtr <- mtr[, unlist(ind)]; 
grp <- grp[colnames(mtr)]; 
```

`r home.url`

# Description

`r WriteDescription(yml$description)`

`r home.url`

# Results

## Summary statistics

```{r missing, include=FALSE}
mss <- length(mtr[is.na(mtr)]); 
if (length(mss==0)) ln <- 'there are no variables including missing values.' else {
  n.mss <- apply(mtr, 1, function(x) length(x[is.na(x)]));
  ln    <- paste('there are', length(n.mss[n.mss>0]), 'variable(s) including missing values.')
};
```

The input data includes:

  - **`r length(unique(grp))`** sample groups
  - **`r ncol(mtr)`** total samples
  - **`r nrow(mtr)`** total variables
    * `r ln`

```{r summary_statistics, include=FALSE}
stat <- cbind(Mean=rowMeans(mtr, na.rm=TRUE), 
              SD=apply(mtr, 1, function(x) sd(x, na.rm=TRUE)), 
              Range=apply(mtr, 1, function(x) diff(range(x, na.rm=TRUE))));
summ <- apply(stat, 2, summary); 
tbl1 <- t(FormatNumeric(summ)); 
```

<div style="color:darkblue">
`r OrderTable()` The mean, standard deviation, and range of all variables.
</div>

`r kable(tbl1) %>% kable_styling(
    bootstrap_options = c("striped", "hover", "condensed", "responsive"), full_width=FALSE)`

`r home.url`

## Variable selection

Run 1-way ANOVA on each variable to identify those significantly different across all sample groups.

```{r anova, include=FALSE}
f <- as.factor(as.vector(grp)); 
dat <- data.frame(t(mtr));
dat$Group <- f;
formula <- as.formula(paste0("cbind(", paste(names(dat)[-length(dat)],collapse=","), ")~Group") ) 
aov <- aov(formula, data=dat);
p <- as.vector(sapply(summary(aov), function(x) x[1, 5])); 
names(p) <- rownames(mtr);

ms <- sapply(unique(grp), function(g) rowMeans(mtr[, grp==g, drop=FALSE], na.rm=TRUE)); 
colnames(ms) <- paste('Mean', colnames(ms), sep='_'); 

stat <- cbind(stat, pANOVA=p, ms); 
saveRDS(stat, paste(path.r, 'stat.rds', sep='/')); 
write.csv(stat, paste(path.tbl, 'stat.csv', sep='/')); 
```

```{r hist_p, include=TRUE, fig.width=6, fig.height=4, out.width='600px'}
# Plot histogram of p values
par(mar=c(5,5,2,2));
hist(p, br=100, col='#E59866', xlab='ANOVA p value', ylab='Number of variables', main='', cex.lab=1.5); 
```

<div style="color:darkblue">
`r OrderFigure()` Distribution of ANOVA p values. Number of variables with p values within each 0.01 interval.
</div>

```{r selection, include=FALSE}
p0 <- sort(p); 
mtr0 <- mtr[names(p0), , drop=FALSE]; 
len <- length(p0[p0<=prms$pvalue]); 
if (len < prms$min) sel <- mtr0[1:min(nrow(mtr0), prms$min), , drop=FALSE] else sel<-mtr0[1:min(len, prms$max), , drop=FALSE];
CreateDatatable(FormatNumeric(stat[rownames(sel), , drop=FALSE]), paste(path.tbl, 'selected.html', sep='/')); 
```

Significant variables were selected using the following criteria:

  - Select variables with ANOVA p values less than **`r prms$pvalue`**
  - Stop if the number of remaining variables is between **`r prms$min`** and **`r prms$max`**, else
    * if the number remaining variable is less than `r prms$min`, select the top `r prms$min` variables with the smallest p values
    * if the number remaining variable is greater than `r prms$max`, select the top `r prms$max` variables with the smallest p values

As a result, **`r nrow(sel)`** variables were selected. Click [here](table/selected.html) to view these variables.

`r home.url`

## Heatmap

```{r heatmap_prepare, include=FALSE}
# function to sort columns
sortColumn<-function(ms, g) {
  corr   <- as.vector(cor(rowMeans(ms[, g[[2]], drop=FALSE]), ms[, g[[1]], drop=FALSE]));
  g[[1]] <- g[[1]][order(corr)];
  for (i in 2:length(g)) {
    corr   <- as.vector(cor(rowMeans(ms[, g[[i-1]], drop=FALSE]), ms[, g[[i]], drop=FALSE]));
    g[[i]] <- g[[i]][rev(order(corr))]; 
  }
  ms[, as.vector(unlist(g))]; 
}

# prepare for heatmap plotting
if (prms$rescale) d <- t(scale(t(sel))) else d <- sel;
d <- d[hclust(as.dist(1-cor(t(d))))$order, , drop=FALSE];
g <- lapply(unique(grp), function(x) names(grp)[grp==x]);
d <- sortColumn(d, g);

w <- max(6, min(1.5+0.25*ncol(d), 10));
h <- min(1.5+0.15*nrow(d), 1.25*w); 
```

```{r heatmap, fig.width=w, fig.height=h, out.width='720px'}
if (prms$rescale) PlotColoredBlock(d, -3, 3, key = 'Normalized value', num.breaks = 127, groups=g) else 
  PlotColoredBlock(d, key = 'Value', num.breaks = 127, groups=g)
```

<div style="color:darkblue">
`r OrderFigure()` Color-coded data of selected variables different across sample groups (red = higher). Variables (rows) were clustered based on their correlation to each other and samples were arranged by groups.
</div>

`r home.url`

***

# Appendix 

Check out the **[RoCA home page](http://zhezhangsh.github.io/RoCA)** for more information.  

## Reproduce this report

To reproduce this report: 

  1. Find the data analysis template you want to use and an example of its pairing YAML file  [here](https://github.com/zhezhangsh/RoCA/wiki/Templates-and-examples) and download the YAML example to your working directory

  2. To generate a new report using your own input data and parameter, edit the following items in the YAML file:

    - _output_        : where you want to put the output files
    - _home_          : the URL if you have a home page for your project
    - _analyst_       : your name
    - _description_   : background information about your project, analysis, etc.
    - _input_         : where are your input data, read instruction for preparing them
    - _parameter_     : parameters for this analysis; read instruction about how to prepare input data

  3. Run the code below within ***R Console*** or ***RStudio***, preferablly with a new R session:

```{r reproduce_report, eval=FALSE, echo=TRUE}
if (!require(devtools)) { install.packages('devtools'); require(devtools); }
if (!require(RCurl)) { install.packages('RCurl'); require(RCurl); }
if (!require(RoCA)) { install_github('zhezhangsh/RoCAR'); require(RoCA); }

CreateReport(filename.yaml);  # filename.yaml is the YAML file you just downloaded and edited for your analysis
```

If there is no complaint, go to the _output_ folder and open the ***index.html*** file to view report. 

## Session information

```{r session_info, echo=FALSE}
sessionInfo(); 
```

`r home.url`

***
_END OF DOCUMENT_
